Geometrization of characters of tori over non-Archimedean local fields

### 2013-08-10 shanghai.tex
Moved the definition of \CS(G(\bFq)) out of the proof of Proposition 10.1

@@

@@

%\begin{remark} \label{rem:CS-action}
The rule $(\gcs{L},\mu,\phi) \mapsto (\gcs{L},\phi)$ defines a forgetful functor from character sheaves on $G$ to $\ell$-adic Weil sheaves on $G$ \cite{Deligne:Weil2}*{Def.~1.1.10 (i)}.
As remarked after \cite{Deligne:Weil2}*{Def.~1.1.10}, a $\ell$-adic Weil sheaf $(\gcs{L},\phi)$ on $G$ may be interpreted as a constructible $\ell$-adic sheaf $\gcs{L}$ on $\bG$ together with an action of the Weil group $\Weil{\Fq}$ on $\gcs{L}$ compatible with the action of $\Gal(\bFq/\Fq)$ on $\bG$. Since we will need this point of view later, we review that interpretation here. Setting 
\[
\phi(\Frob{\Fq}^n) \ceq  \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi)
\]
defines, for every $w\in \Weil{\Fq}$, an isomorphism
\[
\phi(w) : w_G^* \gcs{L}\to \gcs{L},
\]
such that
\[
\phi(uv) = \phi(v) \circ v_G^* \phi(u), \qquad \forall u,v\in \Weil{\Fq}.
\]
Note that $\phi = \phi(\Frob{\Fq})$. 
If we further define $\phi_1(w) \ceq (w_G)_*(\phi(w)^{-1})$ then
\[
\phi_1(w) : (w_G)_* \gcs{L}\to \gcs{L},
\]
is an isomorphism and
\[
\phi_1(uv) = \phi_1(u) \circ (u_G)_* \phi_1(v), \qquad \forall u,v\in \Weil{\Fq},
\]
which means the pair $(\gcs{L},\phi_1)$ is almost an action $\Weil{\Fq}$ on $\gcs{L}$ compatible with the action of $\Gal(\bFq/\Fq)$ in the sense of \cite{SGA4.5}*{Expos\'e XIII,~1.1}. The pair $(\gcs{L},\phi_1)$ fails to fulfill the requirements of \cite{SGA4.5}*{Expos\'e XIII,~1.1} only because $\Weil{\Fq}$ is profinite. This fact goes to the heart of the notion of a Weil sheaf and is one of the essential aspects of our definition of $\CS(G)$.
%\end{remark}

@@

When using character sheaves to study characters, it is useful to understand
how character sheaves behave under change of fields.
Suppose that $k'$ is a finite extension of $k$ of degree $n$, $G$ is an abelian group scheme
over $k$ as in Theorem \ref{thm:snake} and $G_{k'}$ is the base change of $G$ to $k'$.
If $\cs{L} = (\gcs{L}, \mu, \phi)$ is a character sheaf on $G$ we may define
a character sheaf $\nu(\cs{L}) = (\gcs{L}, \mu, \phi_{k'})$ on $G_{k'}$ by setting
\[
\phi_{k'} = \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi).
\]
The commutativity of the diagram (CS.2) for $\phi_{k'}$ follows from the fact that $\Frob{G_{k'}} = \Frob{G}^n$.
Note that we may also think about the construction of $\phi_{k'}$ as taking the action of $\Weil{k}$ defined in Remark
\ref{rem:CS-action} and restricting it to $\Weil{k'}$.

@@

\section{Character sheaves on commutative etale group schemes}

@@

\begin{proposition}\label{prop:etale}
  If $G$ is a commutative \'etale group scheme over $\Fq$ and
  $G(\bFq)$ is finitely generated then there is a canonical
  isomorphism
  \[
  \CSiso{G} \longrightarrow \Hh^1(\Weil{\Fq}, \Hom_\text{grp}(G(\bFq),\EEx)),
  \qquad \text{given by}\qquad
  [\cs{L}] \mapsto [\tau_\cs{L}],
  \]
  where $\tau_\cs{L}: \Weil{\Fq}\to \Hom_\text{grp}(G(\bFq),\EEx)$ is a cocycle such that
  \[
  \forall g\in G(\Fq), \qquad \tau_\cs{L}(\Frob{\Fq})(g) = \trFrob{\cs{L}}(g).
  \]
\end{proposition}

\begin{proof}
  To simplify notation, we set
  $G(\bFq)^* \ceq \Hom_\text{grp}(G(\bFq),\EEx)$.  Recall the
  equivalence between etale group schemes over $\Fq$ and the category
  of groups equipped with a continuous action of $\Gal(\bFq/\Fq)$ (for
  the discrete topology on the group), given by $G \mapsto G(\bFq)$.
  Under this equivalence any character sheaf $\cs{L}=(\gcs{L},\mu,\phi)$
  on $G$ corresponds to:
  \begin{enumerate}
  \labitem{(cs.0)}{cs.0} an indexed set of one-dimensional
    $\EE$-vector spaces $\gcs{L}_x \ceq x^*\gcs{L}$, as $x$ runs over
    $G(\bFq)$;

  \labitem{(cs.1)}{cs.1} an indexed set of isomorphisms
    $\mu_{x,y} \ceq (x \times y)^*\mu : \gcs{L}_{x+y} \mathop{\longrightarrow}\limits^{\iso} \gcs{L}_{x} \otimes \gcs{L}_{y}$,
    for all $x,y \in G(\bFq)$, such that
    \[
    \begin{tikzcd}[row sep=40]
      \gcs{L}_{x+y+z} \arrow{rr}{\mu_{x+y,z}} \arrow[swap]{d}{\mu_{x,y+z}}
      && \gcs{L}_{x+y}\otimes \gcs{L}_{z} \dar{\mu_{x,y} \tight{0.5}{\otimes}{1} \id} \\
      \gcs{L}_{x} \otimes \gcs{L}_{y+z} \arrow{rr}{\id \otimes \mu_{y,z}}
      && \gcs{L}_{x} \otimes\gcs{L}_{y} \otimes \gcs{L}_{z}
    \end{tikzcd}
    \]
    commutes, for all $x,y,z\in G(\bFq)$; and

  \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} \ceq x^*\phi: \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$
    such that
    \[
    \begin{tikzcd}[row sep=40]
      \gcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
      && \gcs{L}_{\Frob{\Fq}(x)}\otimes \gcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
      \gcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
      && \gcs{L}_x \otimes\gcs{L}_y
    \end{tikzcd}
    \]
    commutes, for all $x,y\in G(\bFq)$.
  \end{enumerate}
  Under this equivalence, a map $\varphi : \cs{L} \to \cs{L'}$ of
  character sheaves on $G$ corresponds to:
  \begin{enumerate}
  \labitem{(cs.3)}{cs.3} an indexed set $\varphi_x \ceq x^*\varphi: \gcs{L}_x \to \gcs{L'}_x$
    of linear transformations (of one-dimensional $\EE$-vector spaces),
    as $x$ runs over $G(\bFq)$, such that
    \[
    \begin{tikzcd}[column sep=40]
      \arrow[swap]{d}{\phi_x} \gcs{L}_{\Frob{\Fq}(x)} \rar{\varphi_{\Frob{\Fq}(x)}} & \gcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
      &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}& \arrow[swap]{d}{\mu} \gcs{L}_{x+y} \rar{\varphi_{x+y}} & \gcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
      \gcs{L}_x \rar{\varphi_x} & \gcs{L'}_x
      & {} & \gcs{L}_x\otimes\gcs{L}_y \rar{\varphi_x\otimes \varphi_y} & \gcs{L'}_x \otimes \gcs{L'}_y
    \end{tikzcd}
    \]
    both commute.
  \end{enumerate}
  It is clear that the category defined by conditions \ref{cs.0},
  \ref{cs.1}, \ref{cs.2} and \ref{cs.3}, denoted by
  \cdef{$\CS(G(\bFq))$}, is equivalent to $\CS(G)$. We will prove the
  proposition by exhibiting an isomorphism between
  $\CSiso{G(\bFq)}$ and $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$.

  To produce the isomorphism between $\CSiso{G(\bFq)}$ and
  $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$ we must pick a global section of the line
  bundle $(\gcs{L},\mu)$, for each character sheaf $\cs{L}$ on $G$; 
  such a section exists by Lemma~\ref{lemma:section}.
  We will show that the group isomorphism does not depend on these
  choices.
%
  We use the section $s_\cs{L}$ to define a cocycle $z_\cs{L} \in Z^1(\Weil{\Fq},G(\bFq)^*)$
  as follows. Recall that $\cs{L}$ also determines a family of linear transformations
  $\phi_x : \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$ of one-dimensional $\EE$-vector
  spaces, so $\phi_x(s(\Frob{\Fq}(x)))$ is a scalar multiply of $s(x)$, for
  each $x\in G(\bFq)$; let $\tau_\cs{L}(\Frob{\Fq})(x)$ be that scalar, so
  \begin{equation}\label{t}
    \forall x\in G(\bFq),\qquad \phi_{x}(s(\Frob{\Fq}(x)) = \tau_\cs{L}(\Frob{\Fq})(x) s(x).
  \end{equation}
  Conditions \ref{cs.0} and \ref{cs.2}, together with \eqref{section}, guarantee that
  the scalar $z_\cs{L}(\Frob)(x)$ is non-zero, while condition~\ref{cs.1} forces
  $\tau_\cs{L}(\Frob{\Fq})(x+y) = \tau_\cs{L}(\Frob{\Fq})(x) \ \tau_\cs{L}(\Frob{\Fq})(y)$,
  for all $x,y \in G(\bFq)$.
  % \Clifton{The alternative to this approach is to change the
  %   definition of character sheaves to incorporate the notion of
  %   Galois sheaves. That might not be so bad, but would add another
  %   layer of complication, which we wouldn't use profitably, expect
  %   here.}

>>>>

\section{Notation}

@@

For any group $A$, we will denote by $A^*$ the dual group $\Hom_\text{grp}(A, \EEx)$.

@@

%\begin{remark} \label{rem:CS-action}
The rule $(\gcs{L},\mu,\phi) \mapsto (\gcs{L},\phi)$ defines a forgetful functor from character sheaves on $G$ to $\ell$-adic Weil sheaves on $G$ \cite{Deligne:Weil2}*{Def.~1.1.10 (i)}.

As remarked after \cite{Deligne:Weil2}*{Def.~1.1.10}, a $\ell$-adic Weil sheaf $(\gcs{L},\phi)$ on $G$ may be interpreted as a constructible $\ell$-adic sheaf $\gcs{L}$ on $\bG$ together with an action of the Weil group $\Weil{\Fq}$ on $\gcs{L}$ compatible with the action of $\Gal(\bFq/\Fq)$ on $\bG$. Since we will need this point of view later, we review that interpretation here. Setting 
\[
\phi(\Frob{\Fq}^n) \ceq  \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi)
\]
defines, for every $w\in \Weil{\Fq}$, an isomorphism
\[
\phi(w) : w_G^* \gcs{L}\to \gcs{L},
\]
such that
\[
\phi(uv) = \phi(v) \circ v_G^* \phi(u), \qquad \forall u,v\in \Weil{\Fq}.
\]
Note that $\phi = \phi(\Frob{\Fq})$. 
If we further define $\phi_1(w) \ceq (w_G)_*(\phi(w)^{-1})$ then
\[
\phi_1(w) : (w_G)_* \gcs{L}\to \gcs{L},
\]
is an isomorphism and
\[
\phi_1(uv) = \phi_1(u) \circ (u_G)_* \phi_1(v), \qquad \forall u,v\in \Weil{\Fq},
\]
which means the pair $(\gcs{L},\phi_1)$ is almost an action $\Weil{\Fq}$ on $\gcs{L}$ compatible with the action of $\Gal(\bFq/\Fq)$ in the sense of \cite{SGA4.5}*{Expos\'e XIII,~1.1}. The pair $(\gcs{L},\phi_1)$ fails to fulfill the requirements of \cite{SGA4.5}*{Expos\'e XIII,~1.1} only because $\Weil{\Fq}$ is profinite. This fact goes to the heart of the notion of a Weil sheaf and is one of the essential aspects of our definition of $\CS(G)$. 
%\end{remark}

@@

When using character sheaves to study characters, it is useful to understand
how character sheaves behave under change of fields.
Suppose that $k'$ is a finite extension of $k$ of degree $n$, $G$ is an abelian group scheme
over $k$ as in Theorem \ref{thm:snake} and $G_{k'}$ is the base change of $G$ to $k'$.
If $\cs{L} = (\gcs{L}, \mu, \phi)$ is a character sheaf on $G$ we may define
a character sheaf $\nu(\cs{L}) = (\gcs{L}, \mu, \phi_{k'})$ on $G_{k'}$ by setting
\[
\phi_{k'} = \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi).
\]
The commutativity of the diagram (CS.2) for $\phi_{k'}$ follows from the fact that $\Frob{G_{k'}} = \Frob{G}^n$.
\Clifton{I'll fix this section and link.}
Note that we may also think about the construction of $\phi_{k'}$ as taking the action of $\Weil{k}$ defined in Remark
\ref{rem:CS-action} and restricting it to $\Weil{k'}$.

@@

\section{Character sheaves on commutative \'etale group schemes}

@@

\'Etale group schemes form a counterpoint to connected groups, since the component group
is always \'etale.  The functor $G \mapsto G(\bFq)$ defines an equivalence of categories
between the category of \'etale group schemes over $\Fq$ and the category of groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.

For such a $\Gal(\bFq/\Fq)$-group $G(\bFq)$, we define a category of character sheaves $\CS(G(\bFq))$.
An object in $\CS(G(\bFq))$ is given by the data of

\begin{enumerate}
 \labitem{(cs.0)}{cs.0} an indexed set of one-dimensional
  $\EE$-vector spaces $\gcs{L}_x$, as $x$ runs over
  $G(\bFq)$;

 \labitem{(cs.1)}{cs.1} an indexed set of isomorphisms
  $\mu_{x,y} : \gcs{L}_{x+y} \mathop{\longrightarrow}\limits^{\iso} \gcs{L}_{x} \otimes \gcs{L}_{y}$,
  for all $x,y \in G(\bFq)$, such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{x+y+z} \arrow{rr}{\mu_{x+y,z}} \arrow[swap]{d}{\mu_{x,y+z}}
    && \gcs{L}_{x+y}\otimes \gcs{L}_{z} \dar{\mu_{x,y} \tight{0.5}{\otimes}{1} \id} \\
    \gcs{L}_{x} \otimes \gcs{L}_{y+z} \arrow{rr}{\id \otimes \mu_{y,z}}
    && \gcs{L}_{x} \otimes\gcs{L}_{y} \otimes \gcs{L}_{z}
   \end{tikzcd}
  \]
  commutes, for all $x,y,z\in G(\bFq)$; and

 \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} : \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$
  such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
    && \gcs{L}_{\Frob{\Fq}(x)}\otimes \gcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
    \gcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
    && \gcs{L}_x \otimes\gcs{L}_y
   \end{tikzcd}
  \]
  commutes, for all $x,y\in G(\bFq)$.
\end{enumerate}

A morphism $\varphi : \cs{L} \to \cs{L'}$ is given by 
\begin{enumerate}
 \labitem{(cs.3)}{cs.3} an indexed set $\varphi_x : \gcs{L}_x \to \gcs{L'}_x$
  of linear transformations (of one-dimensional $\EE$-vector spaces),
  as $x$ runs over $G(\bFq)$, such that
  \[
   \begin{tikzcd}[column sep=40]
    \arrow[swap]{d}{\phi_x} \gcs{L}_{\Frob{\Fq}(x)} \rar{\varphi_{\Frob{\Fq}(x)}} & \gcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
    &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}& \arrow[swap]{d}{\mu} \gcs{L}_{x+y} \rar{\varphi_{x+y}} & \gcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
    \gcs{L}_x \rar{\varphi_x} & \gcs{L'}_x
    & {} & \gcs{L}_x\otimes\gcs{L}_y \rar{\varphi_x\otimes \varphi_y} & \gcs{L'}_x \otimes \gcs{L'}_y
   \end{tikzcd}
  \]
  both commute.
\end{enumerate}

The functor $\CS(G) \to \CS(G(\bFq))$ defined by $(\gcs{L}, \mu, \phi) \mapsto (\gcs{L}_x = x^*\gcs{L}, \mu_{x,y} = (x \times y)^* \mu, x^*\phi)$ is easily seen to be an equivalence.  We will use this equivalence implicitly in proving the following proposition.

\begin{proposition}\label{prop:etale}
  If $G$ is a commutative \'etale group scheme over $\Fq$ and
  $G(\bFq)$ is finitely generated then there is a canonical
  isomorphism
  \[
  \CSiso{G} \longrightarrow \Hh^1(\Weil{\Fq}, G(\bFq)^*),
  \qquad \text{given by}\qquad
  [\cs{L}] \mapsto [\tau_\cs{L}],
  \]
  where $\tau_\cs{L}: \Weil{\Fq}\to \Hom_\text{grp}(G(\bFq),\EEx)$ is a cocycle such that
  \[
  \forall g\in G(\Fq), \qquad \tau_\cs{L}(\Frob{\Fq})(g) = \trFrob{\cs{L}}(g).
  \]
\end{proposition}

\begin{proof}
  To produce the isomorphism between $\CSiso{G(\bFq)}$ and
  $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$ we must pick a global section of the line
  bundle $(\gcs{L},\mu)$, for each character sheaf $\cs{L}$ on $G$; 
  such a section exists by Lemma~\ref{lemma:section}.
  We will show that the group isomorphism does not depend on these
  choices.
%
  We use the section $s_\cs{L}$ to define a cocycle $z_\cs{L} \in Z^1(\Weil{\Fq},G(\bFq)^*)$
  as follows. Recall that $\cs{L}$ also determines a family of linear transformations
  $\phi_x : \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$ of one-dimensional $\EE$-vector
  spaces, so $\phi_x(s(\Frob{\Fq}(x)))$ is a scalar multiply of $s(x)$, for
  each $x\in G(\bFq)$; let $\tau_\cs{L}(\Frob{\Fq})(x)$ be that scalar, so
  \begin{equation}\label{t}
    \forall x\in G(\bFq),\qquad \phi_{x}(s(\Frob{\Fq}(x)) = \tau_\cs{L}(\Frob{\Fq})(x) s(x).
  \end{equation}
  Conditions \ref{cs.0} and \ref{cs.2}, together with \eqref{section}, guarantee that
  the scalar $z_\cs{L}(\Frob{k})(x)$ is non-zero, while condition~\ref{cs.1} forces
  $\tau_\cs{L}(\Frob{\Fq})(x+y) = \tau_\cs{L}(\Frob{\Fq})(x) \ \tau_\cs{L}(\Frob{\Fq})(y)$,
  for all $x,y \in G(\bFq)$.
  % \Clifton{The alternative to this approach is to change the
  %   definition of character sheaves to incorporate the notion of
  %   Galois sheaves. That might not be so bad, but would add another
  %   layer of complication, which we wouldn't use profitably, expect
  %   here.}
